Appendix 1: Quantifying Drought

### 2020-11-17 analysis/paper/vital rate models.Rmd
figures for lab meeting talk

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print=FALSE)
library(tidyverse)
library(here)
library(mgcv)
library(dlnm)
library(bbmle)
library(lubridate)
library(glue)
library(patchwork)
library(sf)
library(tsibble)

@@

@@

```{r data, echo=TRUE}
spei <- read_rds(here("analysis", "data", "derived_data", "spei_lags.rds"))
xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
```
# Join data sets

@@

Now join by year and whether plot is in grid cell or not.  Probably need to construct `sf` object for grid cells (knowing that the lat lon are the centers and the resolution is 0.25º) and then use some `sf_` function to check if a point is inside of a grid cell.  See: https://stackoverflow.com/questions/59766153/left-join-based-on-closest-lat-lon-in-r

@@

```{r}
demog_full <-
  demog_full %>% 
  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor))
```

@@

@@

```{r include=FALSE}
#TODO: switch to predict() so y axis is correct
g_cf <-  plot.gam(g_post_cf,  shift = coef(g_post_cf)[1])
g_1ha <- plot.gam(g_post_1ha, shift = coef(g_post_1ha)[1])

g_df <-
  bind_rows(
    "CF" = g_cf[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    "1-ha" = g_1ha[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    .id = "habitat"
  )
```

```{r}
g_df %>% 
  ggplot(aes(x = exp(x), y = exp(fit + coef(g_post_cf)[1]))) +
  geom_jitter(data = demog_full %>% filter(habitat %in% c("1-ha", "CF")),
              aes(x = shts, y = shts_next), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_ribbon(aes(ymin = exp(fit + coef(g_post_cf)[1] - se),
                  ymax = exp(fit + coef(g_post_cf)[1] + se)),
              alpha = 0.2,
              fill = "blue") +
  labs(x = "size (#shoots)", y = "size (t + 1)") +
  facet_wrap(~habitat) +
  theme_bw() +
  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
```
Plants don't grow much.

@@

```{r}
(p_growth_cf | p_growth_1ha) +
  plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
  plot_layout(guides = "collect")
```

@@

```{r}
(p_growth_cf_slice + theme(axis.text.x = element_blank())) /
  p_growth_1ha_slice +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
```

@@

s_df <-
  bind_rows(
    "CF" = s_cf[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    "1-ha" = s_1ha[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    .id = "habitat"
  )
```

@@

```{r}
s_df %>% 
  ggplot(aes(x = exp(x), y = plogis(fit + coef(s_post_cf)[1]))) +
  geom_line() +
  geom_ribbon(aes(ymin = plogis(fit + coef(s_post_cf)[1] - se),
                  ymax = plogis(fit + coef(s_post_cf)[1] + se)),
              alpha = 0.2) +
  labs(x = "size (#shoots)", y = "P(survival)") +
  facet_wrap(~habitat) +
  stat_summary_bin(data = demog_post %>% filter(habitat %in% c("1-ha", "CF")),
                   geom = "point",
                   fun = mean,
                   bins = 20,
                   aes(y = surv, x = shts)) +
  theme_bw() +
  plot_annotation(caption = "Note: these are from two separate models, not a size–habitat interaction!")
```

@@

```{r}
(p_surv_cf|p_surv_1ha) + 
  plot_annotation(title = "Effect of drought on survival") +
  plot_layout(guides = "collect")
```

@@

```{r}
(p_surv_cf_slice / p_surv_1ha_slice) +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, paged.print=FALSE)
library(tidyverse)
library(here)
library(mgcv)
library(dlnm)
library(bbmle)
library(lubridate)
library(glue)
library(patchwork)
library(sf)
library(tsibble)
library(broom)

@@

```{r}
#define plotting themes

theme_talk <- theme_bw() + theme(text = element_text(size = 18))
```

@@

```{r data, echo=TRUE}
spei <- read_rds(here("analysis", "data", "derived_data", "spei_lags.rds"))
xa <- read_rds(here("analysis", "data", "derived_data", "xa_spei_lags.rds"))
demog <- read_rds(here("analysis", "data", "derived_data", "ha_survey.rds"))
plot_coords <- read_csv(here("analysis", "data", "raw_data", "plot_coords.csv"))
```

```{r}
length(unique(demog$ha_id_number))
range(demog$year)
2009-1998
```

# Join data sets

@@

**TODO:**
Now join by year and whether plot is in grid cell or not.  Probably need to construct `sf` object for grid cells (knowing that the lat lon are the centers and the resolution is 0.25º) and then use some `sf_` function to check if a point is inside of a grid cell.  See: https://stackoverflow.com/questions/59766153/left-join-based-on-closest-lat-lon-in-r

@@

```{r}
demog_full <-
  demog_full %>% 
  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
  mutate(flowered = as.numeric(infl > 0))
```

@@

## Flowering Probability

```{r}
f_post_cf <-
  gam(flowered ~ 
        s(log_shts) +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
anova(f_post_cf)
```

```{r}
f_post_1ha <-
  gam(flowered ~ 
        s(log_shts) +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 25),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")
anova(f_post_1ha)
```

```{r}
pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
```

```{r}
flowered_lim <- c(min(pred_f_post_1ha$fitted, pred_f_post_cf$fitted),
                  max(pred_f_post_1ha$fitted, pred_f_post_cf$fitted))
# plogis(flowered_lim)
```

```{r}
p_flowered_cf <-
  pred_f_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
p_flowered_cf
```

```{r}
p_flowered_1ha <-
  pred_f_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous("lag (months)", breaks = 0:24, labels = labs, expand = c(0,0)) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(flowering)", option = "A", limits = plogis(flowered_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
p_flowered_1ha
```

```{r}
p_flowered_cf_slice <-
  pred_f_post_cf %>% 
  filter(nearest(x, -1.50)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "Continuous Forest", y = "P(flowering)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_cf_slice
```

```{r}
p_flowered_1ha_slice <-
  pred_f_post_1ha %>% 
  filter(nearest(x, -1.5)) %>% 
  ggplot(aes(x = lag, y = plogis(fitted))) +
  geom_ribbon(aes(
    ymin = plogis(fitted - se.fit * 1.96),
    ymax = plogis(fitted + se.fit * 1.96)
  ), alpha = 0.3)+
  geom_line() +
  scale_x_continuous("", breaks = 0:24, labels = labs) +
  labs(title = "1-ha Fragments", y = "P(flowering)") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# p_surv_1ha_slice
```

@@

```{r include=FALSE}
Q <- demog_post$spei_history
L <- demog_post$L

newdata_cf <- 
  tibble(
    log_shts = log(
      seq(min(demog_post_cf$shts, na.rm = TRUE),
          max(demog_post_cf$shts, na.rm = TRUE),
          length.out = 50)
    ),
    spei_history = matrix(mean(Q), ncol = ncol(Q), nrow = 50),
    L = L[1:50, ]
  )

newdata_1ha <- 
  tibble(
    log_shts = log(
      seq(min(demog_post_1ha$shts, na.rm = TRUE),
          max(demog_post_1ha$shts, na.rm = TRUE),
          length.out = 50)
    ),
    spei_history = matrix(mean(Q), ncol = ncol(Q), nrow = 50),
    L = L[1:50, ]
  )

g_cf <- augment(g_post_cf, newdata = newdata_cf, type.predict = "link", se_fit = TRUE)
g_1ha <- augment(g_post_1ha, newdata = newdata_1ha, type.predict = "link", se_fit = TRUE)

g_df <-
  bind_rows(
    "CF" = select(g_cf, log_shts, .fitted, .se.fit),
    "1-ha" = select(g_1ha, log_shts, .fitted, .se.fit),
    .id = "habitat"
  )
```

```{r}
p_growth_shoots <-
  ggplot(g_df, aes(x = exp(log_shts), y = exp(.fitted))) +
  geom_jitter(data = demog_full %>% filter(habitat %in% c("1-ha", "CF")),
              aes(x = shts, y = shts_next), alpha = 0.1) +
  geom_line(color = "blue") +
  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96),
                  ymax = exp(.fitted + .se.fit * 1.96)),
              alpha = 0.2,
              fill = "blue") +
  labs(x = "size (#shoots)", y = "size (t + 1)") +
  facet_wrap(~habitat) +
  theme_talk
p_growth_shoots
ggsave(here("growth_shoots.png"), p_growth_shoots, width = 10, height = 5)
```
Plants don't grow much.

@@

```{r}
p_growth_spei <-
  (p_growth_cf | p_growth_1ha) +
  plot_annotation(title = "Effect of drought on growth of average sized plant (~2.8 shoots)") +
  plot_layout(guides = "collect") & theme_talk
p_growth_spei
ggsave(here("growth_spei.png"), p_growth_spei, height = 7, width = 14)
```

@@

```{r}
(p_growth_cf_slice + theme(axis.text.x = element_blank())) /
  p_growth_1ha_slice +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5) on average sized plant (~2.8 shoots)")
ggsave(here("growth slice.png"))
```

@@

s_df <-
  bind_rows(
    "CF" = s_cf[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    "1-ha" = s_1ha[[1]][c("x", "fit", "se")] %>% 
      bind_cols(),
    .id = "habitat"
  )

s_cf <-  augment(s_post_cf, newdata = newdata_cf, type.predict = "link", se_fit = TRUE)
s_1ha <- augment(s_post_1ha, newdata = newdata_1ha, type.predict = "link", se_fit = TRUE)

s_df <-
  bind_rows(
    "CF" = select(s_cf, log_shts, .fitted, .se.fit),
    "1-ha" = select(s_1ha, log_shts, .fitted, .se.fit),
    .id = "habitat"
  )

```

@@

```{r}
p_surv_shoots <-
  ggplot(s_df, aes(x = exp(log_shts), y = plogis(.fitted))) +
  geom_line() +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96),
                  ymax = plogis(.fitted + .se.fit * 1.96)),
              alpha = 0.2) +
  labs(x = "size (#shoots)", y = "P(survival)") +
  facet_wrap(~habitat) +
  stat_summary_bin(data = demog_post %>% filter(habitat %in% c("1-ha", "CF")),
                   geom = "point",
                   fun = mean,
                   bins = 20,
                   aes(y = surv, x = shts)) +
  theme_talk
ggsave(here("surv_shoots.png"), p_surv_shoots, width = 10, height = 5)
```

@@

```{r}
p_surv_spei <- 
  (p_surv_cf|p_surv_1ha) + 
  plot_annotation(title = "Effect of drought on survival") +
  plot_layout(guides = "collect") & theme_talk
ggsave(here("surv_spei.png"), p_surv_spei, width = 14, height = 7)
```

@@

```{r}
((p_surv_cf_slice + theme(axis.text.x = element_blank())) / p_surv_1ha_slice) +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
ggsave(here("survival slice.png"))
```

## Flowering

```{r include=FALSE}
f_cf <-  augment(f_post_cf,  newdata = newdata_cf,  type.predict = "link", se_fit = TRUE)
f_1ha <- augment(f_post_1ha, newdata = newdata_1ha, type.predict = "link", se_fit = TRUE)
f_df <-
  bind_rows(
    "CF" = select(f_cf, log_shts, .fitted, .se.fit),
    "1-ha" = select(f_1ha, log_shts, .fitted, .se.fit),
    .id = "habitat"
  )
```

```{r}
p_flowered_shoots <-
  ggplot(f_df, aes(x = exp(log_shts), y = plogis(.fitted))) +
  facet_wrap(~habitat) +
  geom_line() +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96),
                  ymax = plogis(.fitted + .se.fit * 1.96)),
              alpha = 0.2) +
  labs(x = "size (#shoots)", y = "P(flowering)") +
  stat_summary_bin(data = bind_rows(demog_post_1ha, demog_post_cf),
                   geom = "point",
                   fun = ~mean(., na.rm = TRUE),
                   bins = 20,
                   aes(y = flowered, x = shts)) +
  theme_talk
p_flowered_shoots
ggsave(here("flowering_shoots.png"), p_flowered_shoots, width = 10, height = 5)
```
So something weird is going on with the continuous forest data.  All I know is that if I do everything the same, but take the crossbasis smooth out of the GAM, then it looks right:

```{r}
flowergam <-
  gam(flowered ~ s(log_shts),
      family = binomial,
      data = demog_post_cf,
      gamma = 1.2, #increases smoothness when above 1
      method = "REML")

augment(flowergam, newdata = newdata_cf) %>% 
  ggplot(aes(x = exp(log_shts), y = plogis(.fitted))) +
  geom_line() +
  stat_summary_bin(geom = "point",
                   data = demog_post_cf %>% filter(!is.na(flowered)),
                   aes(x = exp(log_shts), y = flowered),
                   fun = mean,
                   bins = 20) 
```

```{r}
p_flowered_spei <- 
  (p_flowered_cf|p_flowered_1ha) + 
  plot_annotation(title = "Effect of drought on flowering") +
  plot_layout(guides = "collect") & theme_talk
ggsave(here("flowering_spei.png"), p_flowered_spei, width = 14, height = 7)
```

```{r}
((p_flowered_cf_slice + theme(axis.text.x = element_blank())) / p_flowered_1ha_slice) +
  plot_annotation(title = "Effect of severe drought (SPEI = -1.5)")
ggsave(here("flowering slice.png"))
```

### 2020-11-17 analysis/supplementary-materials/appendix_SPEI.Rmd
figures for lab meeting talk

```{r trmm-spei}
library(ncdf4)
nc_open()
```

>>>>

```{r trmm-spei}
# library(ncdf4)
# nc_open()
```

# For Cronians talk
Just gonna make a simple-ish plot (lol) with major droughts labeled

```{r}
xa_df <-
  xa_spei %>% 
  group_by(yearmonth) %>% 
  #Summarize across grid cells
  summarize(`Xavier et al.` = mean(xa_spei))

comb <- left_join(xa_df, gspei %>% 
  mutate(yearmonth = yearmonth(date), date = NULL)) %>% 
  rename(`Global SPEI` = g_spei) %>% 
  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01")))
```

```{r}
p <-
  ggplot(comb %>% filter(dataset == "Xavier et al."), aes(x = yearmonth, y = spei, color = dataset)) +
  geom_line() +
  scale_color_manual(values = c("#3b1255", "darkgreen")) +
  labs(y = "SPEI")
p <- annotate_spei(p)

p <-
  p +
  scale_x_yearmonth(date_breaks = "3 years",
                    date_labels = "%Y") +
  theme_bw() +
  theme(axis.title.x = element_blank(), 
        panel.grid = element_blank(),
        legend.position = "bottom")
p
ggsave(here("spei_xavier.png"), p, width = 10, height = 4, units = "in")
```

Try labeling droughts

```{r}
library(gghighlight)
p + gghighlight(year(yearmonth) ==2010, use_direct_label = FALSE)
```
